---
title: "HW4: MANOVA"
date: "2025-06-14"
output:
  html_document:
    toc: true
    toc_depth: 5
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(MASS)
library(biotools)
library(klaR)
library(car)
library(dplyr)
library(lubridate)
library(ggplot2)
library(ggExtra)
library(heplots)
library(readr)
library(caret)
library(vegan)
```

## Forest Fire Dataset

---

### Dataset Intro & Cleaning 
The [dataset from UCI Machine Learning Respository](link: https://archive.ics.uci.edu/dataset/547/algerian+forest+fires+dataset) includes 244 instances that regroup a data of two regions of Algeria,namely the Bejaia region located in the northeast of Algeria and the Sidi Bel-abbes region located in the northwest of Algeria.

- **region** (categorical): Bejaia or Sidi-Bel Abbès.
- **day** (integer): Day of month
- **month** (categorical): Month of year. From June to September.
- **year** (categorical): Calendar year
- **Temperature** (integer): Temperature at noon (unit: C)
- **RH** (integer): Relative humidity (unit: %)
- **Ws** (integer): Wind speed (unit: km/h)
- **Rain** (continuous): Accumulated rainfall (unit: mm)
- **FFMC** (continuous): Fine Fuel Moisture Code. Measures moisture of surface litter (needles, grass). Higher = drier = more flammable).
  - Depends on temperature, RH, wind, rain, and previous FFMC.
- **DMC** (continuous): Duff Moisture Code. Measures moisture in loosely compacted organic layers (duff) below the surface. Higher DMC -> drier duff -> easier sustained fire.
  - Depends on rain, temperature, RH, and previous DMC
- **DC** (continuous): Drought Code. Measures moisture in deep, compact organic matter. Higher DC -> deeper, long-term drought conditions -> deep-burning fires possible.
  - Depends on  rain, temperature, and previous DC
- **ISI** (continuous): Initial Spread Index. Measures potential fire spread rate immediately after ignition.Higher ISI -> faster spread.
  - Depends on FFMC, wind 
- **BUI** (continuous): Buildup Index. Measures amount of fuel available for combustion. Higher BUI -> more fuel -> more intense fire if ignited.
  - Depends on DMC, DC
- **FWI** (continuous): Fire Weather Index. Measures general fire intensity (combines ISI and BUI). Higher FWI -> higher potential fire severity.
- **Classes** (categorical):; Fire occurrence (`not fire`/`fire`)

#### Data Cleaning
```{r}
df <- read_csv("algerian_forest_fires.csv", skip = 1)
df <- df[complete.cases(df), ]
df$day <- as.integer(df$day)
df$month <- factor(df$month, levels = 6:9, 
                   labels = c("Jun","Jul","Aug","Sep"))
df$year <- as.factor(df$year)
df$DC <- as.numeric(df$DC)
df$FWI <- as.numeric(df$FWI)
df$Classes <- trimws(df$Classes)
df$Classes <- factor(df$Classes)
df$Region <- factor(df$Region)

df$sqWs <- sqrt(df$Ws + 1)
df$tRain <- 1 / ((df$Rain + 1)^3) 
df$tFFMC <- (df$FFMC)^3
df$logDMC <- log(df$DMC + 1)
df$logDC <- log(df$DC + 1)
df$logFWI <- log(df$FWI + 1)
```


In this assignment: 

- Continuous response variables: our three basic meteorological measurements `Temperature`, `RH`, and `Ws`.
- Categorical predictors: `Classes` (fire vs. not fire), `Region` (Bejaia vs. Sidi-Bel Abbes), and `month` (Jun, Jul, Aug, Sep). 
- One or more additional continuous predictors: any other continuous variables in our list. 

---


### Interaction Plots 

#### By categorical predictors `Classes` and `Region`

We start by creating interaction plots for `Temperature`, `RH`, and `Ws`, grouped by the first two categorical predictors `Classes` and `Region`, across all months.

```{r}
resp.vars <- c("Temperature","RH", "Ws")

for (v in resp.vars) {
    interaction.plot(
      x.factor = df$Classes, 
      trace.factor = df$Region, 
      response = df[[v]], 
      type = 'b', lwd = 2, 
      trace.label = "Region",
      lty = c(3, 1), col = c(4, 2), pch = c(16, 4),
      xlab = " ",
      ylab = paste("Mean", v),
      main = paste("Interaction Plot for", v),
    )
  grid() 
}
```

**Observations:**

1. In both regions, temperatures tend to be higher on fire days than on non-fire days. Sidi-Bel Abbes generally records higher temperatures than Bejaia. The lines are nearly parallel, suggesting little to none interaction between Classes and Region.

2. RH is lower on fire days, especially in Sidi-Bel Abbes.
A possible interaction is observed: the RH difference between fire and not fire is greater in Sidi-Bel Abbes.

3. In Sidi-Bel Abbes, mean wind speed is slightly lower on non-fire days than on fire days. 
In Bejaia, it’s the opposite: wind speed is actually higher on non-fire days than on fire days.
This seems a bit counter-intuitive: as it makes more sense for fire risk to increase with higher the wind speeds, the faster the fire spreads. This negative relationship might suggests that wind speed is not a major contributor to fire occurances in the region. 


#### By all 3 ategorical predictors

Out of curiosity, I made interaction plots for the response variables grouped by all three categorical predictors `Classes`, `Region`, and `months`, where Classes = fire is in red, Classes = not fire in blue, Region = Bejaia is in dashed line, and Region = Sidi-Bel Abbes is in solid line. 

```{r}
for (v in resp.vars) {
    interaction.plot(
      x.factor = df$month, 
      trace.factor = interaction(df$Classes, df$Region), 
      response = df[[v]], 
      type = 'b', lwd = 2, 
      trace.label = "Region-Class",
      lty = c(3,3, 1,1), col = c(2, 4, 2, 4), pch = c(16, 16, 4, 4),
      xlab = "Month",
      ylab = paste("Mean", v),
      main = paste("Interaction Plot for", v),
    )
}
```

**Observations:**

1. Temperatures peak in August across all groups. Sidi-Bel Abbes seems to be more fire prone and generally records higher temperatures than Bejaia. For both regions, temperatures tend to be at least one degree higher on fire days than on non-fire days.

2. RH is generally inversely related to fire occurrence: fire groups (red) tend to have lower humidity, especially Sidi-Bel Abbes in August (~45%). 
Bejaia is more humid than Sidi-Bel Abbes. Bejaia—not fire shows the highest RH throughout (~75% in June).
This pattern support the hypothesis that drier air contributes to fire events, especially during midsummer.

  A possible outlier is observed: the RH is higher under fire conditions than non-fire conditions in Bajaia in August, whereas in all other region–class combinations and months, RH consistently drops by approximately 10% from not fire to fire,

3. Wind speed patterns are more variable and region-dependent. Generally, Bejaia seems to be windier than Sidi-Bel Abbes.

---


### Two-Way MANOVA

```{r}
options(contrasts = c("contr.sum","contr.poly")) 
mod_mva <- lm(
  cbind(Temperature, RH, Ws) ~ Classes *Region, data = df
)

summary(Anova(mod_mva, type = 3), univariate = TRUE)
```

**1. Multivariate tests**

  All the tests show highly significant main effects for both `Classes` and `Region` (p-values << 0.001), i.e. fire vs non-fire differ multivariately, so do Sidi-Bel Abbes vs Bejaia.
  However, there is no significant `Classes`×`Region` interaction, i.e. there is no strong evidence that the fire–non-fire difference itself changes between regions.

**2. Univariate follow-ups**

- Fire presence (`Classes`) is significantly associated with higher temperature and lower relative humidity (p-value << 0.01), but not significantly associated with wind speed (p = 0.558).
- `Region` significantly affects all three meteorological variables, i.e. Bejaia and Sidi-Bel Abbes have different climates/microclimates.
- None of the individual responses show a significant Classes×Region term at alpha = 0.05 (RH is borderline at $p\approx 0.067$), matching the non-significant multivariate interaction.

```{r}
cqplot(mod_mva$residuals, label = "Residuals MANOVA")
```

In the chi-square qq plot, most points track the line quite closely except a few pointing a little upwards towards the tail, but this is good enough for us to say that the multivariate‐normality assumption is satisfied. 


#### Three-way MANOVA
Since the two-way MANOVA for Region x Classes is not statistically significant, I decided to add month as a third factor. By doing so, I wanted to ask if the shift in joint distribution of Temperature, RH and Wind attributed to `Classes` and `Region` also change over the seasonal cycle.


```{r}
mod_mva3way <- lm(
  cbind(Temperature, RH, Ws) ~ Classes * Region * month,
  data = df
)

summary(Anova(mod_mva3way, type = 3), univariate = TRUE)
```

**1. Multivariate results**

  When testing Temperature, RH and Wind Speed jointly, all three main effects (`Classes`, `Region` and `month`) are highly significant (each $p << 0.01$), but none of the two-or-three-way interactions reach significance on Pillai’s trace (all $p > 0.05$).
  Roy’s largest-root test for the fire×month term, however, is $p \approx 0.044$, indicating that there is at least one specific linear combination of (T, RH, Ws) in which the fire×month interaction is detectable.
    
  Follow up univariately: in the univariate F‐tests, only Ws has a significant Classes×month effect (F=7.65, p≈0.006), which is the exact response driving Roy’s result.

**2. Univariate follow-ups**

  Looking at each response by itself, Temperature and RH behave as expected: the three response variables each explain a highly significant portion of their variability, but none of their interactions do. 
    
  On the other hand, for wind speed, region remains significant ($p \approx 0.025$), month does not ($p \approx 0.32$), yet the Classes × month F-test for wind has $p \approx 0.006$, which is significant. In other words, in some months the wind speed effect on fire days is larger compared to non-fire days. `Ws` is the exact response driving Roy’s result on the Classes×month effect in the multivariate results.  

---

### Contrasts


```{r}
options(contrasts = c("contr.treatment", "contr.poly"))
contrasts(df$Region) 
contrasts(df$Classes)
```
Therefore, `Region1` is 1 for Sidi‐Bel Abbes (0 for Bejaia), and
`Classes1` is 1 for not‐fire days (0 for fire days).

#### Multivariate contrasts

```{r}
rownames(coef(mod_mva))

# Test 1
linearHypothesis(mod_mva, "Region1 = 0")
```

This tests $H_0: \mu_{B} = \mu_{S}$.
All four multivariate test statistics agree with $p << 0.01$, so we reject \(H_0\) and conclude that the two regions have significantly different joint means of `(Temperature, RH, Ws)`.

```{r}
# Test 2
linearHypothesis(mod_mva, "Classes1 + 0.5*Classes1:Region1 = 0")
```
This tests whether the average fire-effect across both regions is zero, i.e. $$H_0:\ \frac{(\mu_{fire,B} - \mu_{nonfire,B}) + (\mu_{fire,S} - \mu_{nonfire,S})}{2} = 0$$

The resulting p-values are << 0.01, so we reject \(H_0\). Even when averaging the fire vs. non-fire difference across regions, the effect remains highly significant.

```{r}
# Test 3
linearHypothesis(mod_mva,"Classes1 + Classes1:Region1 = 0")
```
This tests whether fire vs. non-fire differ in Sidi-Bel Abbes alone. 
The results suggest that we reject \(H_0\). In Sidi-Bel Abbes, fire days differ significantly from non-fire days.


```{r}
# Test 4
linearHypothesis(mod_mva, "Classes1:Region1 = 0")
```
None of the test results were significant. This confirms again that there is no significant Classes×Region interaction. In other words, fire vs non-fire is comparable in both regions, and the three basic meteorological measurements could be good universal indicators for fire prediction, regardless of region. 



#### Univariate Contrasts

Performing the univariate followup based on the first multivariate contrast (test 1) we have performed: 
```{r}
options(contrasts = c("contr.sum","contr.poly")) 
mod.T <- lm(Temperature ~ Classes * Region, data = df)
mod.RH <- lm(RH ~ Classes * Region, data = df)
mod.Ws <- lm(Ws ~ Classes * Region, data = df)

names(coef(mod.T))

linearHypothesis(mod.T, "Region1 = 0")
linearHypothesis(mod.RH, "Region1 = 0")
linearHypothesis(mod.Ws, "Region1 = 0")
```

The univariate confirm that all three responses differ significantly by region, specially `Temperature` and `RH`. 

```{r}
linearHypothesis(mod.T, "Classes1:Region1 = 0")
linearHypothesis(mod.RH, "Classes1:Region1 = 0")
linearHypothesis(mod.Ws, "Classes1:Region1 = 0")
```

Zooming in to the univariate breakdown of Region x Classes interaction, none of these univariate p-values drops below 0.05. The slight signal in RH ($p\approx 0.067$) matches what we saw before but both temperature and wind show no region‐specific fire effect.


---

### Model with Added Variable
```{r}
pairs(df[c("Rain","Temperature","RH","Ws")],
      pch = 20,
      col = as.numeric(df$Classes),
      main = "logFWI vs. Responses (color = fire class)", 
      panel = function(x, y, ...) {
    points(x, y, ...)
    abline(lm(y ~ x), col = 'grey', lwd = 2)
  }
)
```

In the pair-plots, as `Rain` increases, average temperature tends to drop, relative humidity rises, and wind speed slightly  increase). Considering that rainfall data is always non-negative, and its distribution tend to be naturally highly right skewed, and there isn't too much to do to change that (after several testing), we can loosely accept that the variables are linearly associated.


```{r}
mod_mva2 <- manova(
  cbind(Temperature, RH, Ws) ~ Classes * Region + Rain,
  data = df
)

summary(Anova(mod_mva2, type = 3), univariate = TRUE)

summary.aov(mod_mva2)
```

Multivariate: 

After adjusting for `Rain`, `Classes` and `Region` remain highly significant jointly on `(Temperature, RH, Ws)`. `Rain` also shows a small but significant multivariate effect ($p \approx 0.012$), i.e. adding it to the model explains extra variation across T, RH and Ws.  The Classes × Region interaction stays non-significant.


Univariate: 

- `Temperature` ~ `Rain`: 

  Fire days remain much hotter than non-fire days, S-B Abbes remains cooler than Bejaia, and each extra millimeter of rain cools the day significantly.

- `RH` ~ `Rain`:

  Fires are still drier and S-B Abbes more humid, but rainfall itself no longer shows a clear univariate effect on RH. This suggests that the big humidity contrasts are driven by `Classes` and `Region` rather than by rainfall.

- `Ws` ~ `Rain`:
  
  Wind still does not differ by fire class, but remains higher in S-B Abbes and now also increases modestly with rainfall.


Also note that after we account for rain, the Classes F–stat for Temperature drops from about 78.9 (mod_mva) to 53.3 (mod_mva2), and for RH from about 48.97 to 35.8. This shrinking means part of what we originally attributed to fire effects was actually just days with lower rainfall.

In contrast, the Region F–statistics stay right around 11–12 for Temperature and 35–36 for RH, so the climate difference between Bejaia and Sidi‐Bel Abbes isn’t explained away by rainfall.

---

### Model Assumptions

```{r}
mod_mva2 <- manova(cbind(Temperature, RH, Ws) ~ Classes*Region + Rain, data = df)

cqplot(mod_mva2$residuals, label = "MANOVA residuals")
```

The chi‐square qq plot seems to show reasonably multivariately normal distribution of the residuals. Still, considering the handful of points lying above the upper band, we could try to look at the residuals vs. fitted plots for our dependent variables and perform a box-cox transformation. 

```{r}
mod.T2 <- lm(Temperature ~ Classes * Region + Rain, data = df)
plot(mod.T2, which = c(1,2), pch = 19, col = 'blue')

mod.RH2 <- lm(RH ~ Classes * Region + Rain, data = df)
plot(mod.RH2, which = c(1,2), pch = 19, col = 'blue')

mod.Ws2 <- lm(Ws ~ Classes * Region + Rain, data = df)
plot(mod.Ws2, which = c(1,2), pch = 19, col = 'blue')
```

There is some heteroskedasticity and uneven spread in the Residuals vs. Fitted plot for all there variabels. 

```{r}
bcT  <- powerTransform(mod.T2)
(lamT <- bcT$x[which.max(bcT$y)])

bcH  <- powerTransform(mod.RH2)
(lamH <- bcH$x[which.max(bcH$y)])

bcW  <- powerTransform(mod.Ws2)
(lamW <- bcW$x[which.max(bcW$y)])
```

Interestingly, for each margin, the box–cox log-likelihood is maximized at $\lambda = 1$, i.e. no power transformation. This probably means that the residuals are heavy tailed and aren’t skewed in a way a simple power could fix. 

Moreover, because Pillai’s trace is fairly robust to mild normality violations, the MANOVA conclusions should still remain strong. 

---

### MRPP

```{r}
(mrpp1 <- mrpp(df[,c("Temperature","RH","Ws")], df$Classes))
```

The average within‐group distance in the 3d space of `Temperature`, `RH` and `Ws` was 16.8, lower than the expected value of 18.47 if fire and non-fire days were drawn from the same distribution. The resulting of A is ~0.09 and permutation p-value is 0.001, meaning fire status explains roughly 9% of the total variation (pretty weak) and that the clustering of days by fire vs. non-fire is highly unlikely by chance.


### Conclusion

Meteorological conditions differ meaningfully between fire and non-fire days. Patterns are consistent across regions and months, with temperature and humidity as the strongest indicators. Wind speed seems to show inconsistent associations and contributs less to the multivariate separation. Rainfall adds modest explanatory power but doesn’t override the main effects. Lastly, the overall clustering of fire vs. non-fire days, while statistically significant, was relatively weak. 


Given these limitations, I thought MANOVA did not yield particularly novel insights for this dataset. To further explore MANOVA, I repeated this procedure on the loaner datasets, hoping to see stornger group separation. 

---

## Loaner Dataset: Ohio Crime 

```{r}
crime <- read_csv("ohiocrimehm.csv")
```

Our four dependent variables are:

- GovtJobs (V10): During the summer, the state government should help provide jobs for inner-city youths from poor families.
- RecreationPrograms (V12): Develop recreation programs, like midnight basketball, so that youths will have something to do instead of wandering the streets at night.
- DrugTreatment (V16): Instead of arresting people who use drugs, get them into treatment programs that can help get them off drugs.
- FamilyHelp (V23): Provide help to families and their children as soon as a child shows signs that he or she might later get into trouble with the law.

And 2 predictors (plus their interaction term):

- Gender (V70): 0 = female, 1 = male
- Education (V72): 
  - 1 = never went to high school, 
  - 2 = went to high school but did not graduate, 
  - 3 = graduated from high school,
  - 4 = finished one year of college (or post-high school training), 
  - 5 = finished two years of college, 
  - 6 = finished three years of college, 
  - 7 = graduated from college, 
  - 8 = finished one or more years of graduate school
- Gender x Education interaction

```{r}
responses <- c("V10","V12","V16","V23")
names(responses) <- c(
  "GovtJobs",  
  "RecreationPrograms",
  "DrugTreatment", 
  "FamilyHelp"
)

crime$Gender <- factor(crime$V70, levels=c(0,1), labels=c("Female","Male"))
crime$Race <- factor(crime$V71, levels=c(1,2,3), 
                     labels=c("White","Black","Other"))
crime$Education <-  factor(crime$V72, levels=1:8, 
                           labels=c("noHS", "someHS", "HSGrad",
                                    "College1yr", "College2yr", "College3yr",
                                    "CollegeGrad","GradDegree"))
```

---

### Interaction Plots 
```{r}
for (v in responses) {
  interaction.plot(
    x.factor = crime$Education,
    trace.factor = crime$Gender,
    response = crime[[v]],
    type = "b", lwd = 2, 
    lty = 1, col = c(2, 1, 4), pch = 16:18,
    trace.label = "Race",
    xlab = '',
    ylab = paste("Mean", v),
    main = paste("Interaction Plot for", v),
    cex.axis = 0.6
  )
    grid() 
}

```

For V10, females generally exhibit higher mean scores than males, with both genders following relatively parallel trends across education levels, indicating limited interaction.

In both V12 and V16, the patterns are more irregular, though females still tend to report higher mean scores. Male responses show less variability and consistently lower values, suggesting potential gender differences.

In V23, data for females seem to be limited, but among males, there is strong support for family involvement among those with no high school education. However, this support declines among those with 2–3 years of college education.

Overall, the presence of intersecting lines in these plots suggests interaction effects between gender and education. However, there seem to be a consistent dip in scores at the College3yr level across variables and genders. This may reflect not just genuine attitude differences but also survey-related reasons, for example interpretation ambiguity or response tendencies within this education subgroup.

---

### Two-Way MANOVA
```{r}
options(contrasts = c("contr.sum","contr.poly")) 

crime_mva <- lm(
  cbind(V10, V12, V16, V23) ~ Education *Gender, data = crime
)
```

#### Multivariate Results
```{r}
summary(Anova(crime_mva, type = 3), univariate = TRUE)
```

**Interpretation: ** 

Both `Education` and `Gender` show significant multivariate effect (p < 0.01), confirming that attitudes differ individually among different education levels and the two gender. 

However, the Education × Gender interaction is not significant (p > 0.40), i.e. no evidence that the gender gap changes across education levels. As with the forest fire data, in this dataset, Roy’s test again comes close ($p \approx 0.064$), corresponding to the univariate results where Education × Gender is significant for V10 and V16, but not much from V12 or V23. 

Other univariate followups suggest that for V10 and V12, Education, Gender, and their interactions are highly significant. For V12 and V23, however, no significant effects were found, i.e. consistant responses across groups. 


```{r}
cqplot(crime_mva$residuals, label = "Residuals MANOVA")
```

All data points seem to lie within the bounds, so the residuals are multivariately normally distributed, so assumptions of the model are met. 

---

### Contrasts

```{r}
options(contrasts = c("contr.treatment", "contr.poly"))
contrasts(crime$Gender) 
contrasts(crime$Education)
```

```{r}
rownames(coef(crime_mva))
```

```{r}
linearHypothesis(crime_mva, "Gender1 = 0")
```

This tests overall gender effect. The results shows that there is a significant multivariate gender gap.

```{r}
linearHypothesis(crime_mva, "Education2 = 0")
```

The result shows that high‐school graduation vs. no high‐school shift the joint profile of attitudes in a statistically significant way.

```{r}
linearHypothesis(crime_mva, "Education7 - Education3 = 0")
```

There is a small but statistically significant shift in the combined attitude vector when going from SomeCollege to GradDegree.

```{r}
linearHypothesis(crime_mva, "Gender1 + Education7:Gender1 = 0")
```

This test looks at the gender gap specifically at the top education level. It turns out that the male–female difference among those with a graduate degree is not significant at alpha = .05 despite a weak trend ($p\approx .096$).

```{r}
linearHypothesis(crime_mva, "Education2:Gender1 = 0")
```

Based on the resulting p-values, there is no evidence that the male–female gap at HSGrad differs from the male–female gap at noHS.

```{r}
linearHypothesis(crime_mva, "Education6:Gender1 - Education3:Gender1 = 0")
```

The gender gap at CollegeGrad is not significantly different from the gender gap at HSGrad.

Overall interaction test: 
```{r}
linearHypothesis(crime_mva, c(
  "Education1:Gender1 = 0",
  "Education2:Gender1 = 0",
  "Education3:Gender1 = 0",
  "Education4:Gender1 = 0",
  "Education5:Gender1 = 0",
  "Education6:Gender1 = 0",
  "Education7:Gender1 = 0"
))
```
Jointly, there is no significant interaction. 

Putting everything together, 
Gender has a significant main effect across all four response variables. Big gaps in Education (HSGrad vs noHS) produces significantly different responses,  smaller educaitonal gaps (someHS vs noHS) produces little difference.

Overall, support for crime‐prevention measures changes with gender and to some extent with completing high school, but the gender gap itself does not depend on a person’s education level.

---

### Model with Added Variable


```{r}
crime$Income <- crime$V87

options(contrasts = c("contr.treatment","contr.poly"))

j <- function(x) jitter(x, factor = 1)

panel_jitter <- function(x, y, ...) {
  xi <- j(x); yi <- j(y)
  points(xi, yi, pch = 16, col = rgb(0, 0.5, 1, 0.4), ...)
  ab <- coef(lm(y ~ x))
  abline(ab, col = 2, lwd = 2)
}

pairs(
  crime[c("Income","V10","V12","V16","V23")],
  panel = panel_jitter,
  main = "Income vs. Attitudes (jittered, with fit lines)",
  labels = c("Income","V10","V12","V16","V23")
)

```

Because of the categorical nature of the data, it is hard to tell if the variables are linearly associated. 

```{r}
crime_mod2 <- lm(
  cbind(V10, V12, V16, V23) ~ Gender * Education + Income, data = crime
)

summary(crime_mod2, test = "Pillai")
summary.aov(crime_mod2)
```

For V10 and V16, we see significant negative association with income. Effects from Education and Gender still remain significant, but interaction term is no longer significant.

V12 is very stable and no predictors are statistically significant. 

V23 shows significant difference in responses across genders. 

Overall, adding the income term doesn't change the overall pattern of significance in the multivariate tests, but it explains additional variation, especially in V10 and V16.


---

### Model Assumptions

```{r}
crime_mva2 <- manova(
  cbind(V10, V12, V16, V23) ~ Gender * Education + Income, data = crime)

cqplot(crime_mva2$residuals, label = "MANOVA residuals")

```

Model assumptions are well-met. 

---

### p-value Adjustment 



```{r}
mod_gender <- manova(
  cbind(V10, V12, V16, V23) ~ Gender,
  data = crime
)

summary(mod_gender, test = "Pillai")
aov_gender <- summary.aov(mod_gender)

p_gender <- sapply(
  aov_gender,
  function(x) x["Gender", "Pr(>F)"]
)

p_adj_gender <- p.adjust(p_gender, method = "holm")

gender_contrasts <- data.frame(
  P_value    = p_gender,
  Holm_Adj_P = p_adj_gender
)
print(gender_contrasts)
```

Even after adjusting for multiple comparisons, all four items show a statistically significant gender difference.
In each case, the mean support among women differs from men, so  gender is a robust predictor.
The smallest adjusted p is for V23 (“FamilyHelp”), indicating the strongest gender gap there. 

